---
title: "R for Data Science assessment"
author: "sfhart"
date: "9/22/2022"
output: 
  html_document:
    toc: true
    theme: united
    code_folding: hide
---

<br>

```{r setup, message=FALSE, warning=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
# knitr::opts_chunk$set(echo = TRUE, results = FALSE, message=FALSE, warning=FALSE)
# install.packages("plotly")
# install.packages("tidyverse")
# install.packages("styler")
library(tidyverse)
library(plotly)
library(styler)
library(DT)
```

# Complete the following analysis in R and generate an RMarkdown report to show the analysis and results:

### Read in the gapminder_clean.csv data as a tibble using read_csv

```{r}
# gapminder # note the rows column was unlabeled - X1 for now
gapminder <- read_csv("gapminder_clean.csv") %>%
  as_tibble()
```

<br>

### Filter the data to include only rows where Year is 1962 and then make a scatter plot comparing 'CO2 emissions (metric tons per capita)' and gdpPercap for the filtered data.

```{r}
gapminder1962 <- gapminder %>%
  filter(Year == 1962)
ggplot(gapminder1962, aes(gdpPercap, `CO2 emissions (metric tons per capita)`)) +
  geom_point() +
  xlab("GDP per capita ($)") +
  ylab("CO2 emissions \n (metric tons per capita)") +
  scale_y_log10()+
  scale_x_log10()+
  ggtitle("In 1962 CO2 emissions coorelated with \n GDP/capita")+ 
  theme_classic()+
      theme(plot.title = element_text(hjust = 0.5),
      axis.text=element_text(size=14,face="bold"),
      axis.title=element_text(size=16,face="bold"),
      text=element_text(size=16,face="bold"))    

```

<br>

### On the filtered data, calculate the correlation of 'CO2 emissions (metric tons per capita)' and gdpPercap. What is the correlation and associated p value?

```{r}
# coorlation test
gapminder1962_cor <- cor.test(gapminder1962$`CO2 emissions (metric tons per capita)`, gapminder1962$gdpPercap, use = "complete.obs")
gapminder1962_p <- gapminder1962_cor$p.value %>%
  signif(3)
gapminder1962_c <- gapminder1962_cor$estimate %>%
  round(2)
pearson_output <- paste0("Pearson correlation: ", gapminder1962_c, "\n")
pvalue_output <- paste0("P-value: ", gapminder1962_p)
```
**`r pearson_output`**

**`r pvalue_output`**

**Pearson correlation is an estimate of linear covariance of two continuous variables, appropriate for this comparison of the relationship of CO2 emissions and GDP. The highly significant p-value indicates these values are not independent, which makes intuitive sense that higher GDP countries would release more CO2. Pearson correlation assumes a number of things that could be violated by this data, notably no outliers (you can see one high outlier), which can have a disproportionately high effect on the coorelation. However, in this case it is apparent that the correlation holds even when disregarding this point**

<br>

### On the unfiltered data, answer "In what year is the correlation between 'CO2 emissions (metric tons per capita)' and gdpPercap the strongest?" Filter the dataset to that year for the next step...

```{r}
# Find strongest correlation
  years <- gapminder$Year %>% unique()
  gapminderyear <- gapminder %>%
    group_by(Year) %>%
    summarise(
      coorelation_coefficient = cor.test(`CO2 emissions (metric tons per capita)`, gdpPercap, use = "complete.obs")$estimate
              ) %>%
    arrange(desc(coorelation_coefficient))

  max_year <- gapminderyear[1,1]
  max_year_output <- paste0("\n", "Correlation is strongest in ", max_year)
  gapminder_max_year <- gapminder %>%
    filter(Year == as.numeric(max_year))

# Print out table
  gapminderyear %>%
    datatable(rownames = FALSE,
              colnames = c('Coorelation Coefficient' = 'coorelation_coefficient'),
              caption = htmltools::tags$caption(
                  style = 'caption-side: top; text-align: center;',
                  'Years sorted by coorelation between CO2 emissions and GDP ')
                  )


```
**`r max_year_output`**

**Pearson correlation coefficient is a good proxy for correlation for the same reasons as described in the previous question**


<br>

### Using plotly, create an interactive scatter plot comparing 'CO2 emissions (metric tons per capita)' and gdpPercap, where the point size is determined by pop (population) and the color is determined by the continent. You can easily convert any ggplot plot to a plotly plot using the ggplotly() command.

```{r}
plot1 <- ggplot(gapminder_max_year, aes(x = gdpPercap, y = `CO2 emissions (metric tons per capita)`, size = pop, color = continent)) +
  geom_point() +
  xlab("GDP per capita ($)") +
  ylab("CO2 emissions \n (metric tons per capita)") +
  scale_y_log10()+
  scale_x_log10()+
  ggtitle("CO2 emissions coorelates with GDP/capita")+ 
  theme_classic()+
      theme(plot.title = element_text(hjust = 0.5),
      axis.text=element_text(size=14,face="bold"),
      axis.title=element_text(size=16,face="bold"),
      text=element_text(size=16,face="bold"))    
ggplotly(plot1)
```

<br>
<br>

# Now, without further guidance, use your R Data Science skills (and appropriate statistical tests) to answer the following (use the unfiltered dataset):

### 1. What is the relationship between continent and 'Energy use (kg of oil equivalent per capita)'? (stats test needed)

```{r}
# Filter out NAs
gapminder2 <- gapminder %>%
  filter(
    !is.na("Energy use (kg of oil equivalent per capita)"),
    !is.na(continent)
  ) %>%
  rename(energy = "Energy use (kg of oil equivalent per capita)")

# Run anova (appropriate for quantitative data organized in over 2 categories)
gapminder_aov <- aov(energy ~ continent, data = gapminder2, na.action = na.omit) %>%
  summary()
# gapminder_aov[[1]]
gapminder_aovP <- gapminder_aov[[1]][1, 5]
#gapminder_aovP

# # with log transformation
# gapminder_aov <- aov(log2(energy) ~ continent, data = gapminder2, na.action = na.omit) %>%
#   summary()
# # gapminder_aov[[1]]
# gapminder_aovP <- gapminder_aov[[1]][1, 5]
# gapminder_aovP

# Plot boxplots
ggplot(gapminder2, aes(continent, energy)) +
  # geom_violin() +
  geom_boxplot() +
  ylab("Energy use \n (kg of oil equivalent per capita)")+
  ggtitle("Energy use by continent")+ 
  theme_classic()+
      theme(plot.title = element_text(hjust = 0.5),
      axis.text=element_text(size=14,face="bold"),
      axis.title=element_text(size=16,face="bold"),
      text=element_text(size=16,face="bold"))    


# answer
energy_output <- paste("A: There are big differences in energy use across the 5 continents. This can be seen in the boxplots and is highly significant by ANOVA, p-value:", gapminder_aovP, ". ANOVA, or ANalysis Of VAriance, is appropriate in this case for comparing deviations in means quantitative data organized in over 2 categories. Just like in earlier coorelation, we should be wary about deviations from assumtions, such as the lack of outliers and normal distributios of the data. Even when log-transforming this data to remove outliers and make the data distribution more normal, the ANOVA is highly significiant (even more that before transforming in fact)")
```
**`r energy_output `**

<br>

### 2. Is there a significant difference between Europe and Asia with respect to 'Imports of goods and services (% of GDP)' in the years after 1990? (stats test needed)

```{r}
gapminder1990 <- gapminder %>%
  filter(Year > 1990, (continent == "Asia" | continent == "Europe"))

# Run t-test (appropriate to compare between two categories of quantitative data)
test_output <- t.test(
  filter(gapminder1990, continent == "Asia")$`Imports of goods and services (% of GDP)`,
  filter(gapminder1990, continent == "Europe")$`Imports of goods and services (% of GDP)`
)
ggplot(gapminder1990, aes(continent, `Imports of goods and services (% of GDP)`)) +
  geom_boxplot() +
  ylab("Imports of goods and services \n (% of GDP)") +
  ggtitle("Imports in Asia vs Europe after 1990")+ 
  theme_classic()+
      theme(plot.title = element_text(hjust = 0.5),
      axis.text=element_text(size=14,face="bold"),
      axis.title=element_text(size=16,face="bold"),
      text=element_text(size=16,face="bold"))   


output <- paste("A: There is not a significant difference between the two, as seen in the boxplot and quantified by t.test,  p-value:", test_output$p.value, ". A t.test is appropriate to compare between two categories of quantitative data, and both these country's imports appear normally distributed and thus appropriate for this test.")
```
**`r output`**

<br>

### 3. What is the country (or countries) that has the highest 'Population density (people per sq. km of land area)' across all years? (i.e., which country has the highest average ranking in this category across each time point in the dataset?)

```{r}
### Average and rank for each country ###


gapminder_countries_rank <- gapminder %>%
  group_by(Year) %>%
  mutate(rank = rank(-`Population density (people per sq. km of land area)`,  na.last = "keep")) %>%
  #select(`Country Name`, Year, rank) %>%
  ungroup() %>%
  group_by(`Country Name`) %>%
  summarise(
    avg_density = round(mean(`Population density (people per sq. km of land area)`, na.rm = TRUE)),
    density_avg_rank = round(mean(rank, na.rm = TRUE),1)
            ) %>%
  arrange(desc(avg_density))

gapminder_countries_rank %>%
    datatable(rownames = FALSE,
              colnames = c('Average Density (people/sq.km)' = 'avg_density', 'Average Density Rank' = 'density_avg_rank'),
              caption = htmltools::tags$caption(
                  style = 'caption-side: top; text-align: center;',
                  'Countries sorted by population density')
                  )
  

output <- paste0("A: The highest average across all years is ", gapminder_countries_rank[1, ]$`Country Name`, ". However, the highest average RANKING (rank 1-263, averaged across the 10 years of data, if they exist for the country) is a tie between these two countries: ", gapminder_countries_rank[1, ]$`Country Name`, " and ", gapminder_countries_rank[2, ]$`Country Name`)
```
**`r output`**

<br>

### 4. What country (or countries) has shown the greatest increase in 'Life expectancy at birth, total (years)' since 1962?

```{r}
# List countries and get life exp in 1962 and 2007
  countries <- gapminder$`Country Name` %>% unique()
  gapminder1962 <- gapminder %>%
    filter(Year == 1962, ) %>%
    mutate(`Life expectancy 1962 (years)` = `Life expectancy at birth, total (years)`) %>%
    select(`Country Name`, `Life expectancy 1962 (years)`)
  gapminder2007 <- gapminder %>%
    filter(Year == 2007, ) %>%
    mutate(`Life expectancy 2007 (years)` = `Life expectancy at birth, total (years)`) %>%
    select(`Country Name`, `Life expectancy 2007 (years)`)

# THERE IS PROBABLY A WAY TO DO THIS WITH GROUP_BY WITHOUT HAVING TO MERGE TWO DF, BUT NOT SURE WHAT IT IS
  
# Merge, calculate increase and print table
  gapminder_life_exp <- full_join(gapminder1962, gapminder2007) %>%
    mutate(`Life expectancy increase (years)` = `Life expectancy 2007 (years)` - `Life expectancy 1962 (years)`) %>%
    arrange(desc(`Life expectancy increase (years)`))
  gapminder_life_exp %>%
    datatable(rownames = FALSE,
              caption = htmltools::tags$caption(
                  style = 'caption-side: top; text-align: center;',
                  'Countries sorted by increase in life expectancy')
                  )


  
output <- paste("A:", gapminder_life_exp[1, ]$`Country Name`, "has shown the greatest increase in life expectancy.")
```
**`r output`**
